---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# final.bis620

<!-- badges: start -->
[![R-CMD-check](https://github.com/aespears/final.bis620/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/aespears/final.bis620/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/aespears/final.bis620/branch/main/graph/badge.svg)](https://app.codecov.io/gh/aespears/final.bis620?branch=main)
<!-- badges: end -->

The goal of final.bis620 is to look at the Social Vulnerability dataset and analyze what differentiates regions where the population increases during the day and regions where the population decreases during the day.

## Installation

You can install the development version of final.bis620 from [GitHub](https://github.com/) with:

``` {r}
# install.packages("devtools")
devtools::install_github("aespears/final.bis620")
```

## Example

This plot shows how 

```{r example}
library(final.bis620)
library(sf)
```

```{r}
ct <- fips_data |> dplyr::filter(ST_ABBR == 'CT')
ct$day_change <- day_dif(ct$E_DAYPOP, ct$E_TOTPOP)

plot_map(ct, 'geometry', 'day_change', 'Daytime Population Change in Connecticut')
```

In the analysis described in the vignette, we explore factors that contribute to a change in population like this,
including Per Capita Income and population demographics that might mean people are more likely to commute.
