---
title: "Final Project: Study of Daytime and Total Populations"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{abstract}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup}
library(final.bis620)
library(sf)
library(dplyr)
library(ggplot2)
```

# Background and Motivation
Social vulnerability refers to the potential negative effects a community might face that is caused by external stressors. The CDC/ADSTR Social Vulnerability Index (SCI) uses US census variables in order to identify communities that may need the most support during disasters As part of measuring their Social Vulnerability Index, the CDC gathers information on at the level of census tracts regarding socioeconomic status, household characteristics, racial and ethnic minority status, and housing type and transportation factors.

One included metric that came to my attention when exploring this dataset is the inclusion of the "Daytime population" compared with the total population of a region. It is interesting to note the distinction between daytime and total population. There are many reasons why the population in a region might change throughout the day, primarily relating to daily commute and work patterns - but some regions may experience a increase in population during the day while others may experience a decrease. Many people commute to a city or population center during the day for work but live in a different area due to lower cost of living or other personal/lifestyle preferences.

In my analysis, I wish to explore what factors are most related to an increase in daytime population and what factors are most related to a decrease in daytime population.

# Research Question
My main question is to ask which socioeconomic factors, household characteristics, and geographic factors relate the most to increases and decreases between daytime and full-time population.

# Data Cleaning and Exploration

The first thing that needed to be done is a bit of preliminary data cleaning. In particular, the original dataset uses -999 as a placeholder for missing values, so a function was created to replace them with NAs.
```{r}
df <- fix999(fips_data)
```


Now, because I am most interested in the change between daytime and total population, we get the difference and percent difference between day and total.

```{r}
df$day_change <- day_dif(df$E_DAYPOP, df$E_TOTPOP)
df$pct_day_change <- pct_day_dif(df$E_DAYPOP, df$E_TOTPOP)
```

As one final cleaning step, let's get rid of any rows where this value is missing, since they will not be helpful in our analysis.
```{r}
df <- df |> dplyr::filter(!is.na(df$day_change))
```


Let us visualize these differences now. First we look at the difference, which is
simply the daytime population minus the total population estimate. A positive difference means that there is an increase in population during the day, i.e. people come into the area that do not live there, whereas a negative difference means that there is a decrease in population during the day, i.e people who live in the area leave the area.

```{r}
plot_hist(df, 'day_change', "Daytime Increase in Population")

```

If we just plot the raw change in population during the day, we see many values close to zero and a few very very large increases or decreases. This could be skewed though by areas with high population density and larger populations in general, such as in New York City.


It may be more helpful though to look at the percent increase rather than the raw increase in order to account for regions with higher population density and larger populations in general. 

As we can see in our new histogram, most regions changed by only a small amount, but a few select areas increase by more than 50x or 100x their original population. Even if we filter out only populations in which the percent change in population is more than 100% 
```{r}
df |> dplyr::filter(abs(pct_day_change) < 50) |> 
plot_hist('pct_day_change', "Daytime Percent Increase in Population")
```
If we plot this, it is somewhat hard to interpret due to the skew. 

Let's make a few more plots to explore this relationship further!

First, let's look at a few maps, focusing on Connecticut for now.
```{r}
ct_df <- df |> dplyr::filter(ST_ABBR == 'CT') 
```

```{r}
ct_df |> plot_map('geometry', 'day_change')
```
Just plotting the daytime population change, we see most areas only change a small amount, but it's obscured greatly by the huge increase in daytime population in urban areas like Hartford. For the bulk of the state, it's hard to tell if the population increases or decreases because it all falls within the purple color

Let's try something else. Let's color in our map simply by the binary measure of population increases during the day or decreases during the day.

```{r}
df <- df |> mutate(increase = ifelse(day_change > 0, 'increase', 'decrease'))
ct_df <- ct_df |> dplyr::mutate(increase = ifelse(day_change > 0, 'increase', 'decrease'))
```

```{r}
ct_df |> 
  plot_map('geometry', 'increase', 'Daytime Change in Population')
```
It's interesting to see how these are scattered throughout the state - we don't see as much of a focus around city centers. 

If we plot the region's total population vs the change in population during the day, we see a fairly even spread between increases and decreases across the two total population groups. So even though we see the highest increase during the day in densely populated urban areas like Hartford, it is not the only factor that would contribute to an increase in population during the day.

```{r}
ct_df |> ggplot(aes(x = E_TOTPOP, y = day_change, col = increase)) + geom_point() + theme_bw() + labs(title = "Connecticut Population vs Daytime Change")
```

So now we consider, what factors do play a role in increased or decreased daytime population? Let's make a few observations and visualizations.
First, we do a little bit more data cleaning, as many of the columns in our original data source are redundant as they give percentiles or flags for certain population metrics.
```{r}
df_filtered <- df |> as.data.frame() |> select(ST_ABBR, starts_with('E_'), day_change, pct_day_change, increase)
```


```{r}
facet_hist(df_filtered, cols = c('ST_ABBR', 'increase', 'pct_day_change', 'day_change'),
           title = "Variable histograms") 
```
Almost all of these values seem very far skewed to the right. Let's log transform them to make them more normal for model fitting.

```{r}
df_logtrans <- log_trans(df_filtered, c(starts_with('E_')))
facet_hist(df_logtrans, cols = c('ST_ABBR', 'increase', 'pct_day_change', 'day_change'),
           "Log transformed variable histograms")
```

These look much more symmetric, aside from some spikes at 0. Let's keep them like this for now.




# Analysis

Let us now try and fit a regression model to predict how many people will enter or leave an area during the daytime.

```{r}
model_data <- df_logtrans|> select(-ST_ABBR, -log_E_DAYPOP, -pct_day_change, -increase, -log_E_HH)
summary(lm1 <- lm(day_change ~ ., model_data))
```

We find a lot of these variables are significant! However, our R-squared is quite low, which suggests that the variables account for only a small amount of the variability in daytime population change.   

There are a few factors that could be causing this. My main thought is that there is a great deal of skew in many of our variables because of the proportionally smaller number of FIPS regions with very high populations, as well as how many of our measurements have large spikes at 0.

Future study must take into account the skew, outliers, and heteroscedasticity of these variables with some transformations and other filtering steps in order to get more accurate predictions.Plotting the residuals against the fitted values shows a slightly odd triangular relationship with a few outliers. 
```{r}
resid_plot(lm1)
```



# Interpretation and Conclusion

Now we will interpret these variables in the context of social vulnerability. In particular, we observed significant coefficients relating to household factors as well as to employment. 

One that stands out is a positive coefficient relating to `log_E_PCI`, referring to per capita income. In areas where the residents have higher average income, we see a large influx of population during the day.
This may mean that these areas have too high of a cost of living, and most of those who work in that area live elsewhere where it may be more affordable. 

However, we see negative relationships in areas with higher numbers of people age 65+ or under 17. This may be related to the fact that often retirees or people with young children live in areas outside of urban centers, where the average worker does not live.

Overall, while not definitive, these results provide a bit of preliminary insight to what factors might be contributing to changes in area population during the day. These results are crucial when taking to account factors like transportation, commutes, housing, and zoning. It is important to recognize where people spend their day relative to where they live when making decisions on urban planning. Future analysis should also take into account factors like housing costs, public transport, and urbanization in order to get a more definitive analysis of daytime population.



